\documentclass[twoside]{article}

\usepackage{epsfig}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

\setlength{\oddsidemargin}{0.25 in}
\setlength{\evensidemargin}{-0.25 in}
\setlength{\topmargin}{-0.6 in}
\setlength{\textwidth}{6.5 in}
\setlength{\textheight}{8.5 in}
\setlength{\headsep}{0.75 in}
\setlength{\parindent}{0 in}
\setlength{\parskip}{0.1 in}

\newcommand{\lecture}[3]{
   \pagestyle{myheadings}
   \thispagestyle{plain}
   \newpage
   \setcounter{page}{1}
   \noindent
   \begin{center}
   \framebox{
      \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSE 291A/DSC 291:~Data Systems for Machine Learning, Spring 2026 \hfill} }
       \vspace{6mm}
       \hbox to 6.28in { {\Large \hfill #1  \hfill} }
       \vspace{6mm}
       \hbox to 6.28in{
        \it
        \begin{tabularx}{6.28in}{l>{\raggedleft\arraybackslash}X}
            Lecturer: #2 & Scribes: #3
        \end{tabularx}
       }
      \vspace{2mm}}
   }
   \end{center}
   \markboth{#1}{#1}
   \vspace*{4mm}
}

\begin{document}

\lecture{4: Tensor Formats, MatMul, Accelerators}{Hao Zhang}{
    Ian Jayachandran,
    Samar Karanch,
    Kevin Lin,
    Owen Park,
    Shaurya Raswan,
    Julie Wu,
    Jaewoong Yun
    } % Lecture name, Lecturer, Scribes

\section{Arithmetic Intensity (AI)}

Arithmetic Intensity is a measure of how compute-heavy an operation is relative to how much data needed to fetch from memory to perform that computation.

\textbf{Motivating Example:}

In standard multi-head attention, to get the Q, K, and V matrices, you need to compute each separately ($Q = X @ W_Q$, $K = X @ W_K$, $V= X @ W_V$).

Instead, you can speed up this process by performing a merged QKV by a single matrix operation ($X @ W_{QKV}$ and split into 3 matrices).

\subsection{Arithmetic Intensity Formula:}

\[
    \text{AI} = \text{\# operations } / \text{ \# bytes}
\]

This formula represents the ratio of compute work to memory work. A high AI value means that you are performing a lot of computations for each byte of memory read, while a low AI means the opposite. 

Transformer operations usually have low AI as you need to load entire matrices from memory while performing low-computation like dot products. This means the GPU will just sit idle waiting for loading data.

The practical goal is to raise AI enough so that memory is not a bottleneck in computing speed.

\subsection{Low AI Example 1:}
\begin{lstlisting}[language=C++, title={Simple add() kernel}]
void add(int n, float *A, float *B, float *C) {
    for (int i = 0; i < n; i++) {
        C[i] = A[i] + B[i];
    }
}
\end{lstlisting}

The code computes the operation of $A[i] + B[i]$ $n$ times via the for loop. 

It also loads $A[i], B[i]$ from memory $n$ times each and stores $C[i]$ to memory $n$ times each, so $3n$ computations total.

According to the AI formula, we have $\#$ operations as $n$, and $\#$ memory loads as $3n$. Therefore, $AI = \frac{n}{3n} = \frac{1}{3}$.


\subsection{Low AI Example 2:}

\begin{lstlisting}[language=C++, title={Simple add() and mul() kernel}]
void add(int n, float* A, float* B, float* C){
    for (int i=0; i<n; i++) {
        C[i] = A[i] + B[i];
    }
}

void mul(int n, float* A, float* B, float* C) {
    for (int i=0; i<n; i++) {
        C[i] = A[i] * B[i];
    }
}

float* A, *B, *C, *D, *E, *tmp1, *tmp2;
// assume arrays are allocated here
// compute E = D + ((A + B) * C)
add(n, A, B, tmp1);
mul(n, tmp1, C, tmp2);
add(n, tmp2, D, E);
\end{lstlisting}

Both the add and multiply operations are 1/3 as explained in Example 1. So, if we call the add operation twice and multiply operation once, we have $3n$ total computations and $9n$ total loads and stores. So, $\frac{2n}{6n} = \frac{1}{3}$

\textbf{Therefore, AI is still $\frac{1}{3}$.} 

This is wasteful as we see $tmp1$ being written to memory during add(n, A, B, tmp1), and loaded from memory during mul(n, tmp1, C, tmp2). The same process applies to $tmp2$. Why not keep $tmp1$ or $tmp2$ in memory for quicker processing?

\subsection{More Efficient Example (Kernel Fusion):}


\begin{lstlisting}[language=C++, title={Fused kernel}]
float *A, *B, *C, *D, *E, *tmp1, *tmp2;
// assume arrays are allocated here
// compute E = D + ((A + B) * C)
void fused(int n, float *A, float B, floatC, float *D, float *E) {
    for (int i = 0; i < n; i++) {
        E[i] = D[i] + (A[i] + B[i]) * C[i];
    }
}
// compute E = D + ((A + B) * C)
fused(n, A, B, C, D, E);
\end{lstlisting}

We can instead combine all compute operations in one kernel as shown above. 

Now we have the following:

$\#$ $Operations = 3n$ (2 additions, 1 multiplication for each loop iteration) 

$\#$ $Bytes$ $Loaded = 5n$ (For each iteration, load $A[i], B[i], C[i], D[i]$, and store once to $E[i]$) 

\textbf{AI = 3/5}. Higher AI! Since we lowered the amount of memory accesses, we have sped up our kernel processing time.

\section{Graph Optimization}
\subsection{Parallelization}
\subsection{Parallelization Problems}
- partition
- communicate
- schedule
- consistency
- auto parallelize
\subsection{Runtime \& Scheduling}
- schedule compute / communication / memory as fast as possible
- overlap communication w/ compute
- subject to memory constraints
\subsection{Operator Implementation}
- get fastest possible implementation of MatMul, Conv2d
- optimized for diff GPU types: V100, A100, H100, phone, TPU
deploy model for diff GPU types need adaptation based on varying constraints
- diff precision fp32, fp16, fp8, fp4
- diff shape conv2d\_3x3, conv2d\_5x5, matmul2D, 3D, attention
- goal is to maximize AI

\section{Fast Operators}
\subsection{Vectorization}


\begin{lstlisting}[language=C++, title={Unvectorized}]
float A[256], B[256], C[256]
for (int i = 0; i < 256; ++i) {
    C[i] = A[i] + B[i]
}
\end{lstlisting}


\begin{lstlisting}[language=C++, title={Vectorized}]
for (int i = 0; i < 64; ++i) {
    float4 a = load_float4(A + 1*4);
    float4 b = load_float4(B + i*4);
    float4 c = add_float4(a, b);
    store_float4(C + i* 4, c);
\end{lstlisting}

Vectorization minimizes writes

\subsection{Data Layout}
- store matrix in memory
- stored sequentially (1D, no tensor awareness)
- computation memory limit crucial for quantized ML
- Row Major: A[i, j] = A.data[i * A.shape[1] + j] (left to right)
- Col Major: A[i, j] = A.data[j * A.shape[0] + i] (top to bottom)
- pytorch utilizes row major calc

INCLUDE EXAMPLE PROGRAM ROW/COL MAJOR SETUP
setup row major 2D array, but reading in column major, swap to make program efficient

\subsubsection{Strides}
ML systems store data in strides format: A[i, j] = A.data[offset + i * A.strides[0] + j * A.strides[1]]

offset: offset of tensor relative to underlying storage
strides: strides[i] = \# of elements needed to be skipped in memory to move "one unit" in the i-th dimeniosn of the tensor

\begin{lstlisting}[language=Python]

\end{lstlisting}

A.strides[0] = 1, A.strides[1] = A.shape[0] --> col major
A.strides[0] = A.shape[1], A.strides[1] = 1 --> row major

if a tensor of shape [1,2,3,4] is stored in contiguous in memory following row major, its strides are (24, 12, 4, 1)

strides can separate underlying storage \& view of tensor
enable zero-copy of frequently used operators

\subsubsection{Operator: Slice}
read first 3 rows, 2 columns
change offset by +1, reduce shape to [3,2]
no copy needed of tensor
\subsubsection{Operator: Transpose}
permute by moving first stride to end 
zero copy, underlying storage unchanged
\subsubsection{Operator: Broadcast}
b.strides=[1], b.shape=[1, 3], b.data=[1, 2, 3]
broadcast so b.shape=[4, 3], b.data=[1, 2, 3], b.strides=[?]
add elements to stride tensor
\subsubsection{Operator: Swap Tiles}

\subsubsection{Problems of Strides}
- memory access can become not continuous
- many vectorized ops require continuous storage

\subsection{Parallelization}
elementise, deeper dive will be covered in later lecture
ADD EXAMPLE CODE FROM LECTURE
parallelize through threads
on gpu stream through processors

\section{MatMul}

This section goes over how we can speed up an important operation in neural networks: Matrix Multiplication.

\subsection{MatMul Complexity}
O(N\^3) complexity, but can get O(N\^2.371552)

\subsection{Slow Operation Example of MatMul:} 

\begin{lstlisting}[language=C++]
dram float A[n][n], B[n][n], C[n][n];
for (int i = 0; i < n; ++i) {
    for (int j = 0; j < n; ++j) {
        register float c = 0;
        for (int k = 0; k < n; ++k) {
            register float a = A[i][k];
            register float b = B[j][k];
            c += a * b;
        }
        C[i][j] = c;
    }
}
\end{lstlisting}

This is a standard matrix multiplication that uses 3 registers. The read cost is high since A[i][k] represents one read from DRAM, and B[j][k] represents another read from DRAM. Since we're in a triple nested for-loop, we have $2n^3$ read operations when referencing both matrices. We are also writing to C with a cost of $n^2$.

\subsubsection{How to Make MatMul Fast:}

In order to make matrix multiplication fast, we need to maximize Arithmetic Intensity (keep the number of operations high while decreasing the number of bytes). 

\textbf{How do we do this?} We can do this by reusing the data loaded into the \textbf{registers} and reduce the number of \textbf{DRAM} reads. Because registers are local to processors, this makes registers extremely fast when reading/writing data (essentially 0 latency). DRAM, on the other hand are much slower. However, registers are expensive and small in memory size so we need to find a way to use these efficiently.

To re-use data in registers, we can use approaches like Tiling, which is a technique of re-using data from registers instead of accessing DRAM memory, as shown in the next section.

\subsection{Register Tiled MatMul}
\begin{lstlisting}[language=C++]
dram float A[n/v1][n/v3][v1][v3];
dram float B[n/v2][n/v3][v2][v3];
dram float C[n/v1][n/v2][v1][v2];

for (int i = 0; i < n/v1; ++i) {
    for (int j = 0; j < n/v2; ++j) {
        register float c[v1][v2] = 0;
        for (int k = 0; k < n/v3; ++k) {
            register float a[v1][v3] = A[i][k];
            register float b[v2][v3] = B[j][k];
            c += dot(a, b.T);
        }
        C[i][j] = c;
    }
}
\end{lstlisting}

\textbf{Main Idea: We update our for loop to iterate by block/tile instead of one single element. This reduces read costs since we're reading a tile once and we re-use this across multiple iterations.}

v1, v2, and v3 represent how big your tiles are. For example, if v1 = 2, v2 = 2, v3 = 2, then the tile of A is 2x2, tile of B is 2x2, and output tile of C is 2x2. This is what we load and write into each register for computation.

Recall that in our slower standard MatMul, each iteration loads one element at a time (i.e., register a = A[i][k], where A[i][k] is one element). For our tiled version, \textbf{register "a"} loads a (v1 x v3) tile at a time (a[v1][v3] = A[i][k], where A[i][k] is an entire v1 x v3 block).

We're essentially batching a bunch of numbers and performing dot products for each tiled batch instead of computing one by one. 

Now, lets re-compute the read cost: 

Our total length for our triple-nested for loop is now of size $n/v1 * n/v2 * n/v3$ which simplifies to $n^3 / (v1 * v2 * v3)$. 

Within the loop, reading \textbf{register a} is $v1 * v3$ since we're loading a whole tile of size v1 by v3. So we multiply this by the nested loop size to get $n^2/v2$. The same applies for reading register b.

For write cost, c is still $n^2$ since we still need to store every element regardless of how we tile. So we're speeding up just the read time.

We also need to increase the number of registers needed, since each register can only hold one element. So the number of registers is the sum of our size of our tiles for A (v1 * v3), B (v2 * v3), and C (v1 * v2).

This increase in registers for faster computation is the tradeoff for tiling. The idea is that you want to set v1 and v2 to be as large as possible as long as you have the amount of registers to support it.

\subsubsection{Cache-aware Tiling}
\begin{lstlisting}[language=C++]
dram float A[n/b1][b1][n];
dram float B[n/b2][b2][n];
dram float C[n/b1][n/b2][b1][b2];

for (int i = 0; i < n/b1; ++i) {
    l1cache float a[b1][n] = A[i];
    for (int j = 0; j < n/b2; ++j) {
        l1cache b[b2][n] = B[j];
        
        C[i][j] = dot(a, b.T);
    }
}
\end{lstlisting}

\textbf{Main Idea: Instead of loading tiles into registers, this loads tiles into the L1 cache, which is slower than registers but faster than DRAM.}

You can fit larger tiles in this L1 cache compared to the registers. As shown in the code, we are now loading the entire row of the matrix A and B since the L1 cache is bigger than registers.

We load \textbf{l1cache a} $n/b1$ times via for loop, where each load is size $b1 * n$. So, we have $n^2$ total reads.

We load \textbf{l1cache b} $n/b1 * n/b2$ times via for loop where each load is size $b2*n$. So, we have $n^3 / b1$ total reads.

Choose b1 and b2 small enough so that your tiles can fit in this L1 cache, similar to how v1 and v2 is small enough to fit in registers.



\subsubsection{Combining both Register Tiling and Cache Tiling}

\begin{lstlisting}[language=C++]
dram float A[n/b1][b1/v1][n][v1];
dram float B[n/b2][b2/v2][n][v2];

for (int i = 0; i < n/b1; ++i) {
    l1cache float a[b1/v1][n][v1] = A[i];
    for (int j = 0; j < n/b2; ++j) {
        l1cache b[b2/v2][n][v2] = B[j];
        for (int x = 0; x < b1/v1; ++x) {
            for (int y = 0; y < b2/v2; ++y) {
                register float c[v1][v2] = 0;
                for (int k = 0; k < n; ++k) {
                    register float ar[v1] = a[x][k][:];
                    register float br[v2] = b[y][k][:];
                    C += dot(ar, br.T)
                }
            }
        }
    }
}
\end{lstlisting}

\textbf{Main Idea:} We combine both Register and Cache tiling to minimize data movement among the memory hierarchy. Cache tiling reduces DRAM reads, and register tiling reduces L1 Cache reads. 

For the 2 outer loops i and j, we implement a cache tiling level by loading large tiles of A and B into the L1 cache. The cost of reading \textbf{L1 cache a} is $n^2$, \textbf{L1 cache b} is $n^3/b1$.

Our 3 inner loops x, y, k take small subsets from this L1 cache and put into registers. Then, these registers compute dot products. 

This avoids going back to the DRAM for every multiply by loading into the L1.

\subsubsection{Data Reuse:}
\begin{lstlisting}[language=C++]
float A[n][n];
float B[n][n];
float C[n][n];

C[i][j] = sum(A[i][k] * B[j][k], axis=k)
\end{lstlisting}

Note that the read cost for A depends on v2 and not v1, because A[i][k] is independent of j. This means tiling the j dimension by v2 allows the same tile of A to be reused v2 times, which is exactly where the cost reduction comes from.

\subsection{Homework Candidate}

\end{document}